Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-06-17 assets/responseToReview.txt

-What are the parameters η_0 and η_∞ appeared in Eq. (7)?

Zero-shear viscosity and infinite-shear viscosity as stated in the sentence following eq. (7).

-Please provide more details on the elements of the lattice Boltzmann method, for instance, lattice model considered in this work, the discrete velocities and the equilibrium populations.

Assuming "latticel model" refers to the velocity set: We use D3Q19 (see 4.1), otherwiese please elaborate.
Discrete velocities and equilibrium populations have been added to a new appendix linked from 4.1

@@

-The authors should provide the relation between the relaxation time and viscosity appeared in first term on the right-hand side of Eq. (12). In addition, if Eq. (12) is considered, should the term B_i given by Eq. (9) be modified?

As mentioned below eq (12) this viscosity is used for the LBM solver, therefore the standard LBM relation (see new appendix) applies to this viscosity.
Eq (12) described that \tau is modified, this has implications for eq (9), to reduce ambiguity we intrudced a \prime into the notation and qonsequently Eq (9) has been (slightly) modified and the text below has been expanded for clarity. A reference to \tau^\prime has been introduced below eq (12) as well

### 2024-06-17 NovelLBM.tex
LBM basics appendix

\usepackage[toc, page]{appendix}
\usepackage{float} %better image positions
\DeclareUnicodeCharacter{2212}{-} %Mathplotlib uses 2212 instead of -
\usepackage{adjustbox} %Adjust boxed content
\usepackage{graphicx} %graphics support
% \usepackage{svg} % only works with inkscape shell scirpt as dependency (convert to eps instead) move to pdf
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{sidecap}

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore describe it here only very briefly.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

@@

>>>>

\usepackage[toc, page]{appendix}
\usepackage{float} %better image positions
\DeclareUnicodeCharacter{2212}{-} %Mathplotlib uses 2212 instead of -
\usepackage{adjustbox} %Adjust boxed content
\usepackage{graphicx} %graphics support
% \usepackage{svg} % only works with inkscape shell scirpt as dependency (convert to eps instead) move to pdf
\usepackage{wrapfig}
\usepackage{amsmath}
\setcounter{MaxMatrixCols}{20} % Allow bigger matrices
\usepackage{amssymb}
\usepackage{sidecap}

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore \RR{only mention the key choices here. Further details can be found in the appendix~\ref{app:LBM} or in literature.}
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

@@

	\RR{
	\section{Basics of Lattice-Boltzmann}\label{app:LBM}
	An excellent overview on Lattice-Boltzmann methods (LBM) can be found in the book by Krüger et al. on the subject \cite{Kruger_book}. Our implementation of the base algorithm mostly follows the book, though some implementation details differ. We mention the most important basics of LBM here and recommend the book for further reading.
	Instead of solving the Navier-Stokes equation LBM solves the Boltzmann equation which is a generalization. The Boltzmann equation is discretized both in space $\vec{x}$ and time $t$. Also, the velocity is limited to discrete values $\vec{c}_i$, the so-called velocity sets. The velocity sets are usually denoted as D$d$Q$q$, with $d$ being the number of dimensions and $q$ being the number of discrete velocities. The density is split into the so-called populations $f_i\left(\vec{x}, t\right)$, which are parts of the density moving along a certain discrete velocity at a certain discrete point in time at a certain discrete point in space. With these discretizations the equation reads as follows.
		f_i\left(\vec{x}+\vec{c}_i\Delta t, t+\Delta t\right) = f_i\left(\vec{x}, t\right) + \Omega_i\left(\vec{x}, t\right)
	Here $\Delta t$ is the duration of a time step (distance between the discrete points in time) and $\Omega_i\left(\vec{x}, t\right)$ is the collision operator. The equation describes the populations moving along the discrete velocities to a neighboring point (streaming) while exchanging momentum modeled by the collision operator (collision). The collision operator redistributes the populations, that meet at any given lattice point. This is highly complex and is therefore usually approximated using the Bhatnagar-Gross-Krook (BGK) operator, which reads as follows.
		\Omega_i = - \frac{f_i - f_i^\text{eq}}{\tau_\text{r}}\Delta t
	Here $\tau_\text{r}$ is the relaxation time and $f_i^\text{eq}$ are the equilibrium populations defined as follows.
		f_i^\text{eq} = w_i\rho\left[1 + \frac{\vec{u}\cdot\vec{c_i}}{c_\text{s}^2} + \frac{\left(\vec{u}\cdot\vec{c_i}\right)^2}{2c_\text{s}^4} + \frac{\vec{u}\cdot\vec{u}}{2c_\text{s}^2}\right]
	Here $\vec{u}$ is the local fluid velocity, $\rho$ is the local density, $w_i$ are the lattice weights associated to the chosen velocity set and $c_\text{s}$ is the lattice speed of sound defined as follows.
		c_\text{s}^2 = \frac{1}{3}\frac{\Delta x^2}{\Delta t^2}
	Here $\Delta x$ is the distance between two discrete lattice points.
	The LBM equation with BGK therefore takes the shape of a relaxation towards equilibrium:
		f_i\left(\vec{x}+\vec{c}_i\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right)\right].
	It should be noted, that this is the simple case for a Newtonian fluid. If additional (extern) forces are to be considered, additional terms are added on the right-hand side to account for these. For simple forces we use the algorithm presented by Guo et al. \cite{guo2002}. The following equations relate the discrete quantities back to the physical quantities.
		\rho &= \sum_if_i^\text{eq} = \sum_if_i \\
		\rho\vec{u} &= \sum_i\vec{c}_if_i^\text{eq} = \sum_i\vec{c}_if_i \\
		\frac{\eta}{\rho} &= c_\text{s}^2\left(\tau_\text{r} - \frac{\Delta t}{2}\right)
	Here $\eta$ is the viscosity. These equations are typically solved in lattice units (LU). These result from fixing three values in LU and calculating conversion factors for all other units accordingly from these. The physical values are then divided by these scaling factors. This is mathematically identical to scaling the whole equation and can be used to simplify the calculations and to keep the values in a range optimal for computation. Typically, one picks $\rho_\text{LU} = 1$ and $\eta_\text{LU} = \frac{1}{6}$. Naturally $\Delta x_\text{LU} = 1$ and $\Delta t_\text{LU} = 1$. A convenient third scale is the resolution, meaning the relation of $\Delta x$ to the size of the system. The relaxation time can be freely chosen, but is ideal at unity in LU.
	The LBM scheme with the BGK operator is called single relaxation time (SRT). There are generalizations for two relaxation times (TRT) and for one relaxation time per $i$ (multi relaxation time, MRT). We found the more complex schemes to rarely yield worthwhile improvement and stick to SRT. While there are multiple valid options we use D3Q19 as the velocity set. The 19 velocities and their weights are noted in the following.

	\begin{equation}
		\begin{split}
			&\frac{\Delta t}{\Delta x}\vec{c}_i = \\
			&\begin{Bmatrix}
				0 & 1 & -1 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 \\
				0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & -1 & 1 & 0 & 0 & 1 & -1 \\
				0 & 0 & 0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & -1 & 1 & -1 & 1
			\end{Bmatrix}
		\end{split}
	\end{equation}
	Where each column is a vector.
	\begin{equation}
		w_i = \begin{cases}
			\frac{1}{3},\quad i=0\\
			\frac{1}{18},\quad 1 \leq i \leq 6\\
			\frac{1}{36},\quad \text{otherwise}
		\end{cases}
	\end{equation}
	}
